**Fabryka**

### 2025-05-05 fabryka_pci/article.Rmd
Suggestion for various typos and biblio improvements

You can use Fabryka $\beta$ 3.0 by following this link
<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
updated to Fabryka 1.0 after the review process of this article. The
Fabryka application files, made in R [@rcoreteam2025] by using *shiny* package [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak

@@

Fabric is made of the orientation and inclination of elongated
sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It is
a statistic (*i.e.*, it needs as much measurements as possible) that
enables the identification of sediment sources that make up the deposits
[@texier2020], sedimentary processes involved in their formation
[@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be considered as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, the fabric analysis
made of the sedimentary deposits depends mainly on the archaeological
material (bones and lithics), which enables them to be inserted into the
formation dynamics of the site and to obtain information on their
representativeness for the reconstruction of past activities.

@@

As proposed by McPherron (2018), a spatial investigation of the fabric
is proposed in a dedicated panel. Benn indices (1994) calculated for
each spatial series of remains are projected within a Benn diagram in
which each pole (isotropic, linear, and planar) is coloured respectively
in red, green and blue. The RGB colour code is reused to materialize the
fabric of the remains on spatial projections (plan and section views) of the material along the axes of the excavation grid. \bigbreak

@@

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a dataset coming from
Le Moustier Lower shelter (Dordogne, France). The example files directly
contain the good column names. You can use them directly in the
application. Whatever the names of the columns in your file, you can
select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(*i.e.* no empty modalities, so no line without a fabric measurement).
When a data table with new columns appears below the column selector,
you are ready to use the application.

@@

-   Benn diagrams with possible integration of Bertran and Lenoble (2002) model (fig. \ref{fig:figure_classical_method}D)
-   Rose diagrams for bearing and plunge (fig. \ref{fig:figure_classical_method}G)
-   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
-   Summary data table with Benn indices, additional statistics, and orientation statistical tests.

@@

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues (see Woodcock 1977). They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axis of the space. They are used to compute
    the Benn indices
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles (Krumbein 1939).

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters ('filter by sample'
and 'filter by code') and the slider (*i.e.* minimum number of measures) in the left
sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
Certain geological processes can have a different impact on objects
depending on their physical properties (*e.g.* silex versus bones in a
water flow as in the example files provided where bones have a strong
preferential orientation and are much more 'linear' in the Benn diagram,
see [@thomas2019] and [@texier2020]). The natural processes models that
you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel. In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations (Curray 1956) in the
summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analyses, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. rose diagrams subpanel, H. Schmidt diagram subpanel, I. summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

@@

-   Benn diagrams with Benn indices computed with nearest neighbours points (McPherron 2018)
-   Spatial projections of the points according to their fabric shape
-   Summary data tables with orientation statistical tests, one at the
    scale of each spatial series (each object and its n nearest
    neighbours fabric = one row) and one at the scale of the sample (fig.
    \ref{fig:figure_spatialised_method}F and G).

@@

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues [@woodcock1977]. They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axes of the space. They are used to compute
    the Benn indices.
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles (see Krumbein 1939).

@@

# Util links

-   use Fabryka $\beta$ 3.0:
    <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source codes: <https://github.com/marchaeologist/fabryka/tree/main>
    \smallbreak
-   github reprository doi of Fabryka $\beta$ 3.0: \smallbreak

>>>>

You can use Fabryka $\beta$ 3.0 by following this link
<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
updated to Fabryka 1.0 after the review process of this article. The
Fabryka application files, made in R [@rcoreteam2025] by using the R package *shiny* [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak

@@

Fabric is made of the orientation and inclination of elongated
sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It is
a statistic (*i.e.*, it needs as much measurements as possible) that
enables the identification of sediment sources that make up the deposits
[@texier2020], sedimentary processes involved in their formation
[@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be treated as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, the fabric analysis
made of the sedimentary deposits depends mainly on the archaeological
material (bones and lithics), which enables them to be inserted into the
formation dynamics of the site and to obtain information on their
representativeness for the reconstruction of past activities.

@@

As proposed by @mcpherron2018, a spatial investigation of the fabric
is proposed in a dedicated panel. Benn indices (1994) calculated for
each spatial series of remains are projected within a Benn diagram in
which each pole (isotropic, linear, and planar) is coloured respectively
in red, green and blue. The RGB colour code is reused to materialize the
fabric of the remains on spatial projections (plan and section views) of the material along the axes of the excavation grid. \bigbreak

@@

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a dataset coming from
Le Moustier Lower shelter (Dordogne, France). The example files directly
contain the right column names. You can use them directly in the
application. Whatever the names of the columns in your file, you can
select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(*i.e.* no empty modalities, so no line without a fabric measurement).
When a data table with new columns appears below the column selector,
you are ready to use the application.

@@

-   Benn diagrams with possible integration of @bertran2002 model (fig. \ref{fig:figure_classical_method}D)
-   Rose diagrams for bearing and plunge (fig. \ref{fig:figure_classical_method}G)
-   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
-   Summary data table with Benn indices, additional statistics, and orientation statistical tests.

@@

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues (see Woodcock 1977). They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axis of the space. They are used to compute
    the Benn indices
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles [@krumbein1939].

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters ('filter by sample'
and 'filter by code') and the slider (*i.e.* minimum number of measures) in the left
sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
Certain geological processes can have a different impact on objects
depending on their physical properties (*e.g.* silex versus bones in a
water flow as in the example files provided where bones have a strong
preferential orientation and are much more 'linear' in the Benn diagram,
see @thomas2019 and @texier2020). The natural processes models that
you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel. In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations [@curray1956] in the
summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analyses, D. buld your model thanks to @bertran2002 data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. rose diagrams subpanel, H. Schmidt diagram subpanel, I. summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

@@

-   Benn diagrams with Benn indices computed with nearest neighbours points [@mcpherron2018]
-   Spatial projections of the points according to their fabric shape
-   Summary data tables with orientation statistical tests, one at the
    scale of each spatial series (each object and its n nearest
    neighbours fabric = one row) and one at the scale of the sample (fig.
    \ref{fig:figure_spatialised_method}F and G).

@@

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues [@woodcock1977]. They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axes of the space. They are used to compute
    the Benn indices.
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues [@benn1994]. These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles [@krumbein1939].

@@

# Useful links

-   use Fabryka $\beta$ 3.0:
    <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source code: <https://github.com/marchaeologist/fabryka/tree/main>
    \smallbreak
-   Zenodo repository of Fabryka $\beta$ 3.0: <https://doi.org/10.5281/zenodo.15207419> \smallbreak
